An exposition on the propriety of restricted Boltzmann machines

### 2016-09-19 writing/draft.Rmd
updated to block means and add table

```{r models-load}
source("helpers/functions.R")
load("data/sample_images.Rdata")

@@

```{r M_eff, cache=TRUE}
###
### MBB Statistic
###
mbb_mean <- function(data_star, data, b) {
  m <- length(data_star)
  n <- length(data)
  N <- n - b + 1

@@

  sqrt(m)*(mean(data_star) - mean(blockmeans))
}

###
### MBB Sample
###
mbb_sample <- function(data, b) {
  n <- length(data)
  k <- floor(n/b)

  start <- sample.int(n - b + 1, k, replace = TRUE)
  data_star <- NULL
  for(i in 1:k) {
    data_star <- c(data_star, data[start[i]:(start[i] + b - 1)])
  }

  return(data_star)
}

###
### Block Bootstrap function
###
bb <- function(x, b, B, boot_stat, boot_sample) {
  #ii. For blocksize b, generate B MBB versions of statistic
  bb.stat <- array(NA, B)
  for(i in seq_len(B)) {
    bb.stat[i] <- boot_stat(boot_sample(x, b), x, b)
  }
  return(bb.stat)
}

marginal_sample_good %>% 
  group_by(v1, v2, v3, v4, image_id) %>%
  do(mbb = bb(.$prob, 2000^(1/3), 2000, mbb_mean, mbb_sample)) -> marginal_mbb

trunc_sample_good %>% 
  group_by(v1, v2, v3, v4, image_id) %>%
  do(mbb = bb(.$prob, 5000^(1/3), 2000, mbb_mean, mbb_sample)) -> trunc_mbb

marginal_mbb %>%
  group_by(v1, v2, v3, v4, image_id) %>%
  do(data.frame(C = var(.$mbb[[1]]))) %>%
  left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
  mutate(M_eff = sigma2/C) %>%
  ungroup() %>%
  summarise(M_eff = min(M_eff)) %>% as.numeric() -> marginal_M_eff

trunc_mbb %>%
  group_by(v1, v2, v3, v4, image_id) %>%
  do(data.frame(C = var(.$mbb[[1]]))) %>%
  left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
  mutate(M_eff = sigma2/C) %>%
  ungroup() %>%
  summarise(M_eff = min(M_eff)) %>% as.numeric() -> trunc_M_eff
```

Secondly, we can assess the mixing of our chain using an idea of effective sample size. If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of image $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior probability of image $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of image $i$ and $M$ is the length of the chain. However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$. We use a block bootstrap method with overlapping blocks to get an estimate for $C_i$, $\hat{C}_i$, and compare it to an estimate of $\sigma^2_i$ using the raw chain, $\hat{\sigma}^2_i$. Formally,

@@

Using this metric, a chain of length $M = 1000$ for the marginalized likelihood method yields $\min_i M_{eff}^{(i)} = `r round(1000*marginal_M_eff, 2)`$, while the truncated Normal method only yields $\min_i M_{eff}^{(i)} = `r round(1000*trunc_M_eff, 2)`$. Thus, the truncated likelihood would require $`r round(marginal_M_eff/trunc_M_eff, 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution. For this reason, in conjuction with the ACF results, the truncated Normal method is not an effective method for fitting the RBM.

>>>>

```{r models-load}
load("data/sample_images.Rdata")

@@

```{r M_eff, cache=TRUE}
###
### Overlapping block means
###
overlap_mean <- function(data, b) {
  n <- length(data)
  N <- n - b + 1

@@

  blockmeans
}

marginal_sample_good %>% 
  group_by(v1, v2, v3, v4, image_id) %>%
  do(means = overlap_mean(.$prob, 2000^(1/3))) -> marginal_means

trunc_sample_good %>% 
  group_by(v1, v2, v3, v4, image_id) %>%
  do(means = overlap_mean(.$prob, 5000^(1/3))) -> trunc_means

marginal_means %>%
  group_by(v1, v2, v3, v4, image_id) %>%
  do(data.frame(C = 2000^(1/3)*var(.$means[[1]]))) %>%
  left_join(marginal_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
  mutate(M_eff = sigma2/C,
         model = "Marginal Likelihood") %>%
  bind_rows(trunc_means %>%
    group_by(v1, v2, v3, v4, image_id) %>%
    do(data.frame(C = 5000^(1/3)*var(.$means[[1]]))) %>%
    left_join(trunc_sample_good %>% group_by(v1, v2, v3, v4) %>% summarise(sigma2 = var(prob))) %>%
    mutate(M_eff = sigma2/C,
           model = "Truncated Normal Prior")) %>%
  select(-C, -sigma2) %>%
  spread(model, M_eff) %>%
  ungroup() %>%
  select(-(v1:v4)) %>%
  gather(Model, `$M_{eff}$`, -image_id) %>%
  mutate(`$M_{eff}$` = 1000*`$M_{eff}$`) %>%
  spread(image_id, `$M_{eff}$`) -> M_eff_table

M_eff_ratio <- M_eff_table[1,-1]/M_eff_table[2,-1]
```

Secondly, we can assess the mixing of our chain using an idea of effective sample size. If the MCMC chain were truly iid draws from the target distribution, then the estimate of the posterior mean probability of image $i$, $\bar{p}^{(i)}$ would be distributed approximately Normal with mean the true posterior probability of image $i$, $p^{(i)}$, and variance $\sigma^2_i/M$, where $\sigma^2_i$ is the true posterior variance of image $i$ and $M$ is the length of the chain. However, with the presence of correlation in our chain, the aysmptotic variance of $\bar{p}^{(i)}$ is instead some $C_i/M$, where $C_i$ is some positive constant such that $C_i > \sigma^2_i$. We can use an overlapping blockmeans approach to get a crude estimate for $C_i$, $\hat{C}_i = b \text{Var} \overline{p_i}$ where $b$ is the block length and $\overline{p_i}$ are the blockmeans for image $i$. We compare it to an estimate of $\sigma^2_i$ using the raw chain, $\hat{\sigma}^2_i$. Formally,

@@

```{r m_eff_table}
M_eff_table %>%
  gather(Image, M_eff, -Model) %>%
  spread(Model, M_eff) %>%
  mutate(Image = as.numeric(Image)) %>%
  arrange(Image) -> M_eff_tmp

cbind(M_eff_tmp[1:8,], M_eff_tmp[9:16,]) %>%
  kable(digits = 2, caption = paste0("\\label{tab:m_eff}The effective sample sizes for a chain of length $M = 1000$ of all $16$ images. The truncated likelihood would require at least $", round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1), "$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution."))
```

The effective sample sizes for a chain of length $M = 1000$ of all $16$ images are presented in table \ref{tab:m_eff}. These range from  $`r round(min(M_eff_table[1, -1]), 2)`$ to $`r round(max(M_eff_table[1, -1]), 2)`$ for the marginalized likelihood method, while the truncated Normal method only yields between $`r round(min(M_eff_table[2, -1]), 2)`$ and $`r round(max(M_eff_table[2, -1]), 2)`$ effective draws. Thus, the truncated likelihood would require at least $`r round(min(M_eff_table[1, -1])/min(M_eff_table[2, -1]), 1)`$ times more iterations to be run of the MCMC chain in order to achieve the same amount of effective information about the posterior distribution. For this reason, in conjuction with the ACF results, the truncated Normal method is not an effective method for fitting the RBM.
